On Verification Patterns for Low-Level Systems via Modal Abstractions

### 2025-09-11 paper.tex
+

\paragraph{Hybrid Logic} Hybrid logics~\cite{blackburn1995hybrid,goranko1996hierarchies,gargov1993modal,areces2001hybrid,brauner2010hybrid} are a class of modal logics that add two primary new concepts to a logic. Nominals $\ell \in \textsf{Loc}$ uniquely identify points in a model (e.g., a particular state), so there is exactly one point in the model where a nominal $\loc$ is true. Satisfaction operators are modal operators indexed by nominals, which enable claims about the truth of another proposition at some arbitrary point in the model identified by a nominal. Traditionally $@_{\ell}\varphi$ asserts that $\varphi$ is true in the state identified by $\ell$. Conceptually, given our context, we might like $l$ to indicate truth in a particular data structure $l$.  However, most prior work on hybrid logics works in classical logics rather than a substructural setting like separation logic. 
Newer work surveyed here adapts this idea to deal with different structures specified in
separation logic.
%We adapt these ideas to assertions of the form $@[l]\varphi$ asserts that $\varphi$ is true in a data structure with \emph{logical} state $\sigma$ \emph{at location} $l$.
% In practice we must design variants of this for each kind of structure we wish to apply this to, e.g. tree nodes, free list, or snapshots.
%This section provides details of and models for each modal abstraction, focusing on how each data structure can act as a model for a data-structure-specific logic.
%We continue to motivate the development with aspects of the write amplification process described in the previous section; Section \ref{sec:evaluation} provides further examples using the modal abstractions in proofs.

@@

An example use of state naming explicitly in the assertion appears in program logics 
with usage protocols (e.g. state transition systems resembling typestate~\cite{strom1986typestate,garcia2014foundations}, which ensure
the operations applied to some data occur in a specified order, and that invariants for various conceptual states of the data
--- such as a file being open or closed --- are respected). 
Protocol assertions are \emph{annotated with the name of the last (abstract) state at which the protocol is ensured}.
A classic application of this idea to systems is 
Halpern et al.'s work adapting modal logics of
knowledge to deal with distributed systems~\cite{halpern1990knowledge,halpern1989modelling,halpern2017reasoning}.
In most of that line of work, $\mathcal{K}_a(P)$ indicates that the node $a$ in the system \emph{knows} or \emph{possesses knowledge of} $P$
(for example, a Raft node may ``know'' a lower bound on the commit index).
Alternatively, a modality $@_i(P)$ may represent that $P$ is true of/at the specific node $i$~\cite{gordon2019modal} (e.g., that node $i$ has stored a certain
piece of data to reliable storage).
These permit capturing specific concepts relevant to the correctness (and reasoning about correctness) of a certain class of systems,
involving facts about specific named entities in the system.
\\

In each of these cases, the fact that these facts are described using modalities with the core modal property $P \Rightarrow Q$, then $M(P) \Rightarrow M(Q)$ means, for example, that if a
process $p$ knows that the commit index is greater than 5 (e.g., $\mathcal{K}_p(\mathsf{commitIndex} > 5)$)
no extra work is required to conclude that the node knows it is greater than 3 (i.e., that $\mathcal{K}_p(\mathsf{commitIndex} > 3)$),
because this follows from standard properties of modal operators as described above.
In contrast, if verification instead used a custom assertion \[\mathsf{minCommitIndex}(5) \] to represent the former knowledge,
one would need to separately provide custom reasoning to conclude \[\mathsf{minCommitIndex}(3)\].
As our survey shows, the value of the modal view (which grants both intuitive specifications and natural
reasoning principles) is becoming apparent in systems software verification.

@@

\subsubsection{Virtualization of Memory Locations}
\label{sec:backgroundonmachinemodel}
A typical general purpose computer virtualizes  memory resources in RAM.
A program asking for a memory unit from an operating system kernel is served with a memory address that is \emph{virtualized}
such that it may be \emph{mapped to} a physical RAM location or not. 
As our informal definition suggests, the common technique for \emph{virtualization} of memory locations in OS 
kernels relies \emph{translating} (\emph{mapping}) virtual addresses into physical ones. 
The conceptual address-translation map is implemented with processor's page-table trees as shown in 
Figure \ref{fig:pagetables}. In this Figure, we see a typical 4 levels of page tables, a virtual address on the 
top indexing into the different levels of the page-table tree (the larger rectangles along the bottom
are blocks of memory which serve as nodes of the tree along a particular lookup path). 
Fabrication of virtual addresses out of limited physical addresses is provided with entries 
in the page-table tree where each points to a physical address aligned to 4KB (4096 byte) boundaries. 
An address translation requires traversal of a series of tables starting from level 4, and it is 
conventionally referred as a \emph{page-table walk}. 
The traversal ends in the level 1 page table with the final lookup to the actual page of physical memory holding the requested data, and the low-order 12 bits being used to index into this page.
\begin{figure}
    \includegraphics[width=\columnwidth]{pagetables2.png}
    \caption{x86-64 page table lookups.}
    \label{fig:pagetables}
\end{figure}

@@

\subsubsection{Virtualized Resources in Filesystems}
Another important computer resource computer resource that is \emph{virtualized} for access is disk blocks. 
Filesystems are software components that abstract disk access on behalf of normal programs
Similar to how VMMs using page-table trees to fabricate more memory addresses than what actually exists, 
filesystems uses indexing maps for fabricating more \emph{virtual disk-page} addresses which are conventionally called \emph{disk-pages} 
to be distinguished from physical \emph{disk-block} addresses.
When a regular program accesses a file, it requests a specific logical offset from the start of that file's contents.
As with virtual memory, the filesystem may map adjacent logical blocks of a single file to physically disparate parts of the disk (or even
to different disks), though this translation mapping is implemented entirely in software (as opposed to the hardware cooperation that exists
for memory virtualization).

>>>>

\paragraph{Hybrid Logic} Hybrid logics~\cite{blackburn1995hybrid,goranko1996hierarchies,gargov1993modal,areces2001hybrid,brauner2010hybrid} are a class of modal logics that add two primary new concepts to a logic. Nominals $\ell \in \textsf{Loc}$ uniquely identify points in a model (e.g., a particular state), so there is exactly one point in the model where a nominal $\loc$ is true. Satisfaction operators are modal operators indexed by nominals, which enable claims about the truth of another proposition at some arbitrary point in the model identified by a nominal. Traditionally $@_{\ell}\varphi$ asserts that $\varphi$ is true in the state identified by $\ell$. Conceptually, given our context, we might like $l$ to indicate truth in a particular data structure $\ell$.  However, most prior work on hybrid logics works in classical logics rather than a substructural setting like separation logic. 
Newer work surveyed here adapts this idea to deal with different structures specified in
separation logic.
%We adapt these ideas to assertions of the form $@[l]\varphi$ asserts that $\varphi$ is true in a data structure with \emph{logical} state $\sigma$ \emph{at location} $l$.
% In practice we must design variants of this for each kind of structure we wish to apply this to, e.g. tree nodes, free list, or snapshots.
%This section provides details of and models for each modal abstraction, focusing on how each data structure can act as a model for a data-structure-specific logic.
%We continue to motivate the development with aspects of the write amplification process described in the previous section; Section \ref{sec:evaluation} provides further examples using the modal abstractions in proofs.

@@

An example use of state naming explicitly in the assertion appears in program logics 
with usage protocols (e.g. state transition systems resembling typestate~\cite{strom1986typestate,garcia2014foundations}, which ensure
the operations applied to some data occur in a specified order, and that invariants for various conceptual states of the data
--- such as a file being open or closed --- are respected). 
Protocol assertions are \emph{annotated with the name of the last (abstract) state at which the protocol is ensured}.
A classic application of this idea to systems is 
Halpern et al.'s work adapting modal logics of
knowledge to deal with distributed systems~\cite{halpern1990knowledge,halpern1989modelling,halpern2017reasoning}.
In most of that line of work, $\mathcal{K}_a(P)$ indicates that the node $a$ in the system \emph{knows} or \emph{possesses knowledge of} $P$
(for example, a Raft node may ``know'' a lower bound on the commit index).
Alternatively, a modality $@_i(P)$ may represent that $P$ is true of/at the specific node $i$~\cite{gordon2019modal} (e.g., that node $i$ has physically stored a certain
piece of data to reliable storage).
These permit capturing specific concepts relevant to the correctness (and reasoning about correctness) of a certain class of systems,
involving facts about specific named entities in the system.
\\

In each of these cases, the fact that these facts are described using modalities with the core modal property that if $P \Rightarrow Q$, then $M(P) \Rightarrow M(Q)$. This means, for example, that if a
process $p$ knows that the commit index is greater than 5 (e.g., $\mathcal{K}_p(\mathsf{commitIndex} > 5)$)
no extra work is required to conclude that the node knows it is greater than 3 (i.e., that $\mathcal{K}_p(\mathsf{commitIndex} > 3)$),
because this follows from standard properties of modal operators as described above.
In contrast, if verification instead used a custom assertion \[\mathsf{minCommitIndex}(5) \] to represent the former knowledge,
one would need to separately provide custom reasoning to conclude \[\mathsf{minCommitIndex}(3)\].
As our survey shows, the value of the modal view (which grants both intuitive specifications and natural
reasoning principles) is becoming apparent in systems software verification.

@@

\subsubsection{Virtualization of Memory Locations}
\label{sec:backgroundonmachinemodel}
A typical general purpose computer virtualizes  memory resources in RAM.
A program asking for a memory unit from an operating system kernel is served with a memory address that is \emph{virtualized}
such that it may be \emph{mapped to} a physical RAM location or not. 
As our informal definition suggests, the common technique for \emph{virtualization} of memory locations in OS 
kernels relies \emph{translating} (\emph{mapping}) virtual addresses into physical ones. 
The conceptual address-translation map is implemented with processor's page table trees as shown in 
Figure \ref{fig:pagetables}. In this Figure, we see a typical 4 levels of page tables, a virtual address on the 
top indexing into the different levels of the page table tree (the larger rectangles along the bottom
are blocks of memory which serve as nodes of the tree along a particular lookup path). 
Fabrication of virtual addresses out of limited physical addresses is provided with entries 
in the page table tree where each points to a physical address aligned to 4KB (4096 byte) boundaries. 
An address translation requires traversal of a series of tables starting from level 4, and it is 
conventionally referred as a \emph{page table walk}. 
The traversal ends in the level 1 page table with the final lookup to the actual page of physical memory holding the requested data, and the low-order 12 bits being used to index into this page.
\begin{figure}
    \includegraphics[width=\columnwidth]{pagetables2.png}
    \caption{x86-64 page table lookups.}
    \label{fig:pagetables}
\end{figure}

@@

\subsubsection{Virtualized Resources in Filesystems}
Another important computer resource computer resource that is \emph{virtualized} for access is disk blocks. 
Filesystems are software components that abstract disk access on behalf of normal programs
Similar to how VMMs using page table trees to fabricate more memory addresses than what actually exists, 
filesystems uses indexing maps for fabricating more \emph{virtual disk-page} addresses which are conventionally called \emph{disk-pages} 
to be distinguished from physical \emph{disk-block} addresses.
When a regular program accesses a file, it requests a specific logical offset from the start of that file's contents.
As with virtual memory, the filesystem may map adjacent logical blocks of a single file to physically disparate parts of the disk (or even
to different disks), though this translation mapping is implemented entirely in software (as opposed to the hardware cooperation that exists
for memory virtualization).
